---
title: "Visualization of benchmarking results"
output:
  html_document:
    df_print: paged
---

## load necessary libraries

```{r,message=FALSE}
library(Seurat)
library(ggplot2)
library(circlize)
library(ComplexHeatmap)
library(grid)
library(cowplot)
library(ggpubr)
library(dplyr)
source("functions.R")

set.seed(1)
```

## input data
```{r}
#location of folder with Seurat objects of test datasets
seurat_objects_dir = "~/projects/demux-benchmarking/results/all_datasets/seurat/"
benchmarking_results_dir = "~/projects/demux-benchmarking/manuscript/results/"
```

## Effect of ambient contamination on demultiplexing
```{r,message=FALSE}
dataset = "CMO_nuclei"
assay = "HTO"
seurat_object = readRDS(sprintf("%s/%s.rds",seurat_objects_dir,dataset))

seurat_object = runUMAP(seurat_object, normalization_margin = 1)
seurat_object$ground_truth = factor(seurat_object$ground_truth)

mtrx = GetAssayData(seurat_object,assay = assay ,slot = "counts") %>% as.matrix()
df = mtrx %>% reshape2::melt()
colnames(df) = c("Tag", "cell_id","counts")
#df$cell_type = sapply(df$Tag, FUN = function(x) { barcode2cellType[[x]] })

#df$Tag = sapply(df$Tag, FUN = function(x) {paste0(x ," (" , barcode2cellType[[x]] , ")") })
df$Tag = as.factor(df$Tag )
df$Tag  = factor(df$Tag , levels = stringr::str_sort(levels(df$Tag ), numeric = T) )

vln_raw = ggplot(df, aes(x=Tag, y=log1p(counts), fill=Tag)) +
  geom_violin() +theme_classic()+
  theme(legend.position="right")+
  theme(text = element_text(size=13), plot.title =element_text(face = "bold"),
        axis.text.x = element_text(angle=-45, hjust=0,vjust = 1,  size = 13))+
  scale_fill_manual(values = c('#F68282','#28CECA','#B95FBB','gray'))+
   ylab("log (raw_counts + 1)") + ggtitle("A    Before Normalization")

##
seurat_object = seurat_object %>% NormalizeData(normalization.method = "CLR",margin = 1,
                                                assay = "HTO")

FeaturePlot(seurat_object, features = rownames(seurat_object)[1:2], max.cutoff = "q95",min.cutoff = "q5", slot = "data", keep.scale = "all")

###### 
raw_counts = GetAssayData(seurat_object,assay = assay ,slot = "counts") %>% as.matrix() %>% t() %>% as.data.frame() 
topHTO <- apply(raw_counts, 1, function(x) names(x)[x == max(x)][[1]])

norm_counts = GetAssayData(seurat_object,assay = assay ,slot = "data") %>% as.matrix() %>% t() %>% as.data.frame() 
topHTO_norm <- apply(norm_counts, 1, function(x) names(x)[x == max(x)][[1]])

seurat_object$topHTO = topHTO
seurat_object$topHTO_norm = topHTO_norm

seurat_object$topHTO = factor(seurat_object$topHTO)


my_cols <- c('DU145'='#F68282','MCF7'='#28CECA','MDAMB231'=  '#B95FBB','PC3'='gray', 'Doublet' = "black")
topTag_rawCounts = DimPlot(seurat_object, group.by = "topHTO", cols = my_cols,pt.size = 0.1)+ggtitle("D      Top Tag (raw counts)")  + theme(plot.title = element_text(hjust = 0))

topTag_normCounts = DimPlot(seurat_object, group.by = "topHTO_norm", cols = my_cols,pt.size = 0.1)+ggtitle("E      Top Tag (Normalized counts)") + theme(plot.title = element_text(hjust = 0))

ground_truth_umap = DimPlot(seurat_object, group.by = "ground_truth", cols = my_cols, pt.size = 0.1)+ggtitle("C      Ground Truth") + theme(plot.title = element_text(hjust = -0))

table(topHTO)
##############
mtrx = GetAssayData(seurat_object,assay = assay ,slot = "data") %>% as.matrix()
df = mtrx %>% reshape2::melt()
colnames(df) = c("Tag", "cell_id","Normalized_counts")
df$Tag = as.factor(df$Tag )
df$Tag  = factor(df$Tag , levels = stringr::str_sort(levels(df$Tag ), numeric = T) )

vln_norm = ggplot(df, aes(x=Tag, y=Normalized_counts, fill=Tag)) +
  geom_violin() +theme_classic()+
  theme(legend.position="right")+
  theme(text = element_text(size=13),axis.text.x = element_text(angle=-45, hjust=0,vjust = 1,  size = 13) , plot.title =element_text(face = "bold") )+
  scale_fill_manual(values = c('#F68282','#28CECA','#B95FBB','gray'))+
  ylab("Normalized counts") + ggtitle("B    After per-tag CLR Normalization")

( vln_raw + vln_norm)/ ( topTag_rawCounts + topTag_normCounts + ground_truth_umap)


figure1 <- ggpubr::ggarrange(plotlist = list((vln_raw + vln_norm)/ (ground_truth_umap + topTag_rawCounts + topTag_normCounts) ),ncol = 1, nrow = 1)
figure1
#pdf("~/projects/demux-benchmarking/manuscript/Figures/fig_1A.pdf", width = 14, height = 8)
#figure1
#dev.off()
```

## Effect of sequence depth variability on demultiplexing
```{r}
dataset = "LMO_custom_cells"
tag = "DU145"

seurat_object = readRDS(sprintf("%s/%s.rds",seurat_objects_dir,dataset))

seurat_object$total_counts = log(seurat_object$nCount_HTO)
seurat_object = seurat_object %>% runUMAP(normalization_margin =1)


my_cols <- c('DU145'='#F68282','MCF7'='#28CECA','MDAMB231'=  '#B95FBB','PC3'='gray', 'Doublet' = "black")

p1 = DimPlot(seurat_object, group.by = "ground_truth",cols = my_cols)+ ggtitle("") + scale_color_discrete(name = "Ground truth") 
p2 = FeaturePlot(seurat_object,features = "total_counts",max.cutoff = "q95", min.cutoff = "q5")+ ggtitle("") + scale_color_continuous(name = "Total tag counts", type = "viridis") 

seurat_object = seurat_object %>% runUMAP(normalization_margin = 2)
p3 = DimPlot(seurat_object, group.by = "ground_truth",cols = my_cols)+ ggtitle("") + scale_color_discrete(name = "Ground truth") 
p4 = FeaturePlot(seurat_object,features = "total_counts",max.cutoff = "q95", min.cutoff = "q5")+ ggtitle("")+ scale_color_continuous(name = "Total tag counts", type = "viridis") 


figure <- ggpubr::ggarrange(plotlist = list(p1 ,p3, p2,p4),ncol = 2, nrow = 2,hjust = 0, labels = c("A","B","C","D"),label.x = 0.001)
figure2 = annotate_figure(figure,top = text_grob("No sequencing depth normalization                                            With sequencing depth normalization", rot = 0, size = 15,face = "bold"))

figure2
# pdf("~/projects/demux-benchmarking/manuscript/Figures/fig_1B.pdf", width = 13, height = 10)
# figure2
# dev.off()
```

## separability analysis

```{r,message=FALSE}

df = read.csv(sprintf("%s/separability_analysis.csv",benchmarking_results_dir),header = T)

ggplot(df, aes(x = reorder(Normalization , -AUCPR), y = AUCPR, fill =Normalization)) + 
  geom_violin() + labs(title="",x="Demuxplexing workflows", y = "aucPR")+ 
  geom_jitter(color="black", size=0.4, width = 0.1) +
  theme_cowplot()+ theme(text = element_text(size=13),
                         axis.text.x = element_text(angle=-45, hjust = 0,vjust=1, size = 14)) 


#####
df2 = reshape2::dcast(df, Dataset + tag ~ Normalization, value.var = "AUCPR")
df2$label = ifelse(df2$raw < 0.8, df2$Dataset,"")

title_color = "red"
title_size = 14

p1 = ggplot(data = df2,aes(y = CLR_1, x = raw)) +
  geom_point(aes()) + xlab("AUCPR of raw counts")+
  #ggrepel::geom_text_repel(seed = 1, size=4)+
  ylab("AUCPR of normalized counts")+ 
  xlim(c(0.55,1))+ylim(c(0.55,1))+
  theme(text = element_text(size=12),legend.position = "None",plot.title = element_text(size = title_size, color = title_color),
                         axis.text.x = element_text(angle=0, hjust=0,vjust = 1,  size = 12),
                         axis.text.y = element_text( size = 12),
                         axis.title.x =  element_text( size = 11),
                         axis.title.y =  element_text( size = 11)) +
  geom_abline(linetype = "dashed") + ggtitle("CLR_exp_bias")

p2 = ggplot(data = df2,aes(y = CLR_2, x = raw)) +
  geom_point(aes()) + xlab("AUCPR of raw counts")+
  #ggrepel::geom_text_repel(seed = 1, size=4)+
  ylab("AUCPR of normalized counts")+ 
  xlim(c(0.55,1))+ylim(c(0.55,1))+
  theme(text = element_text(size=12),legend.position = "None",plot.title = element_text(size = title_size,color = title_color),
                         axis.text.x = element_text(angle=0, hjust=0,vjust = 1,  size = 12),
                         axis.text.y = element_text( size = 12),
                         axis.title.x =  element_text( size = 11),
                         axis.title.y =  element_text( size = 11)) +
  geom_abline(linetype = "dashed")+ggtitle("CLR_seq_depth")

p3 = ggplot(data = df2,aes(y = RC, x = raw, label = label)) +
  geom_point(aes()) + xlab("AUCPR of raw counts")+
  ggrepel::geom_text_repel(seed = 1, size=4)+
  ylab("AUCPR of normalized counts")+ 
  xlim(c(0.55,1))+ylim(c(0.55,1))+
  theme(text = element_text(size=12),legend.position = "None", plot.title = element_text(size = title_size,color = title_color),
                         axis.text.x = element_text(angle=0, hjust=0,vjust = 1,  size = 12),
                         axis.text.y = element_text( size = 12),
                         axis.title.x =  element_text( size = 11),
                         axis.title.y =  element_text( size = 11)) +
  geom_abline(linetype = "dashed")+ggtitle("Relative counts/LogNomralize",)


p1 + p2 +p3
```

## examples of increased separability after normalization

```{r,message=FALSE}
dataset = "lung_cell_line"
seurat_object = readRDS(sprintf("%s/%s.rds",seurat_objects_dir,dataset))

tag = rownames(seurat_object)[3]
title_color = "red"
title_size = 14


seurat_object2 = subset(seurat_object, subset = ground_truth != "Doublet")
seurat_object2$pos_neg= ifelse(grepl(pattern = tag,x = seurat_object2$ground_truth ),paste0(tag,"+ cells"),paste0(tag,"- cells"))
res = separability_analysis(seurat_object2, tag = tag,positive_cells = "pos_neg", norm.methods = c("CLR_2","RC","CLR_1","LogNormalize"))

plots = list()
for(p in names(res$histograms)){
  grob <- grobTree(textGrob(sprintf("AUCPR = %0.2f",res$AUCPR[[p]]), x=0.4,  y=0.90, hjust=0,
                            gp=gpar(col="black", fontsize=11))) 
  plots[[p]] = res$histograms[[p]] + annotation_custom(grob) + theme_classic()
}

raw = plots$raw + ggtitle("Raw counts") + theme( plot.title = element_text(size = title_size,color = title_color))
clr1 = plots$CLR_1 + ggtitle("CLR_exp_bias ")+ theme( plot.title = element_text(size = title_size,color = title_color))
clr2 =  plots$CLR_2 +ggtitle("CLR_seq_depth")+ theme( plot.title = element_text(size = title_size,color = title_color))
RC =  plots$RC +ggtitle("Relative counts")+ theme( plot.title = element_text(size = title_size,color = title_color))
logNorm = plots$LogNormalize +ggtitle("LogNormalize")+ theme( plot.title = element_text(size = title_size,color = title_color))
(raw + clr1 + clr2) / (logNorm + RC)


seurat_object = seurat_object %>% NormalizeData(normalization.method = "CLR",margin = 1,
                                                assay = "HTO")
seurat_object$dataset = dataset
RidgePlot(seurat_object, features = tag, group.by = "dataset" , log = T, slot = "counts", ncol = )|
  RidgePlot(seurat_object, features = tag , group.by = "dataset" , log = F, slot = "data", ncol = 1)
```


## Figure #1

```{r}
figure <- ggpubr::ggarrange(plotlist = list(p1+p2+p3,raw+clr1+clr2+logNorm+RC),ncol = 1, nrow = 2,hjust = 0, labels = c("A","B"),label.x = 0.001)

figure
# pdf("~/projects/demux-benchmarking/manuscript/Figures/fig_1.pdf", width = 12, height = 8)
# figure
# dev.off()


```

## Comparison of normalization-based workflows with existing tools (real datasets)

```{r,message=FALSE}
file = sprintf("%s/real_datasets.csv",benchmarking_results_dir)
df = read.table(file,header = F)
colnames(df) = c("Dataset","Demultiplexing_method","tag", "metric", "value")
df = df[df$metric == "Fscore",]


qc = read.csv(sprintf("%s/sample_level_metrics.csv",benchmarking_results_dir), header = T, row.names = 1)

top_workflows = df %>% dplyr::group_by(Demultiplexing_method) %>% dplyr::summarise(avg_fscore = mean(value)) %>% dplyr::arrange(-avg_fscore)

df$Demultiplexing_method = factor(df$Demultiplexing_method, levels = top_workflows$Demultiplexing_method)

mtrx = df %>% reshape2::dcast(formula = paste0("Demultiplexing_method", "~", "Dataset"),
                                value.var = "value")
rownames(mtrx) = mtrx[,"Demultiplexing_method"]
mtrx = mtrx[,c(2:ncol(mtrx))]
mtrx = mtrx %>% as.matrix()
qc = qc[colnames(mtrx),]
col_fun = circlize::colorRamp2(breaks = c(0,0.8,1),
                               hcl_palette = "BlueRed 2", reverse = F)
col_fun2 = colorRamp2(c(min(qc$seq_depth_var), mean(qc$seq_depth_var), max(qc$seq_depth_var)), c('gray', "white", "brown"))
column_ha = HeatmapAnnotation("Sequencing depth variability" = qc$seq_depth_var,col = list("Sequencing depth variability" = col_fun2), annotation_legend_param = list("Sequencing depth variability" = list(direction = "horizontal")))

ht = ComplexHeatmap::Heatmap(mtrx %>% as.matrix(), col =col_fun, cluster_rows = F, cluster_columns = T,
                        row_names_gp = gpar(fontsize = 13) ,top_annotation=column_ha,
                        heatmap_legend_param = list(direction = "horizontal"),
                        column_names_gp = gpar(fontsize = 13) ,
                        name = "Average F-score",column_names_rot = -45,
                        cell_fun = function(j, i, x, y, width, height, fill) 
                        {grid.text(sprintf("%0.2f", mtrx[i, j]), x, y, gp = gpar(fontsize = 13,
                                                                                 col = "black"))},
                        row_names_side = "right", column_names_side = "bottom" )

Fscore_heatmap = draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")

Fscore_heatmap
```
## Doublet calling
```{r}

file = sprintf("%s/doublet_calling.txt",benchmarking_results_dir)

df2 = read.table(file,header = F)
colnames(df2) = c("Dataset",	"Demultiplexing_method",	"classification",	"proportion")

df2$Demultiplexing_method = as.factor(df2$Demultiplexing_method)
df2$Demultiplexing_method = factor(df2$Demultiplexing_method, levels = top_workflows$Demultiplexing_method)

Dblt.as.snglts = df2[df2$classification == "Singlet", ]

mtrx = Dblt.as.snglts %>% reshape2::dcast(formula = paste0("Demultiplexing_method", "~", "Dataset"),value.var = "proportion")

rownames(mtrx) = mtrx[,"Demultiplexing_method"]
mtrx = mtrx[,c(2:ncol(mtrx))]
mtrx = mtrx %>% as.matrix()
qc = qc[colnames(mtrx),]
col_fun = circlize::colorRamp2(breaks = c(0,0.5 , 0.8),
                               hcl_palette = "BlueRed 2", reverse = F)
col_fun2 = colorRamp2(c(min(qc$seq_depth_var), mean(qc$seq_depth_var), max(qc$seq_depth_var)), c('gray', "white", "brown"))
column_ha = HeatmapAnnotation("Sequencing depth variability" = qc$seq_depth_var,col = list("Sequencing depth variability" = col_fun2), annotation_legend_param = list("Sequencing depth variability" = list(direction = "horizontal")))

ht = ComplexHeatmap::Heatmap(mtrx %>% as.matrix(), col =col_fun, cluster_rows = F, cluster_columns = T,row_names_gp = gpar(fontsize = 13) ,top_annotation=column_ha,
                        heatmap_legend_param = list(direction = "horizontal"),
                        column_names_gp = gpar(fontsize = 13) ,
                        name = "Proportion of doublets predicted as signlets", column_names_rot = -45,
                        cell_fun = function(j, i, x, y, width, height, fill) 
                        {grid.text(sprintf("%0.2f", mtrx[i, j]), x, y, gp = gpar(fontsize = 13,
                                                                                 col = "black"))},
                        row_names_side = "right", column_names_side = "bottom" )

dblt_heatmap = draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dblt_heatmap 
```

## LMO_custom_cells example
```{r}
dataset = "LMO_custom_cells"
seurat_object = readRDS(sprintf("%s/%s.rds",seurat_objects_dir,dataset))
tag = rownames(seurat_object)[4]

totalCellsPerTag = data.frame(ground_truth = seurat_object$ground_truth) %>%
dplyr::group_by(ground_truth) %>% dplyr::summarise(total = n())  

predictions = read.table(sprintf("%s/lmo_custom_predictions.txt",benchmarking_results_dir), sep = "\t", header = T)

df = inner_join(totalCellsPerTag, predictions)
df$percentage = df$num_cells / df$total
df = df[df$ground_truth ==tag ,]

df$predicted_label[df$predicted_label != df$ground_truth & df$predicted_label %in% rownames(seurat_object)] = "Other singlets"
df =df[df$predicted_label != df$ground_truth, ]

df=df %>% dplyr::group_by(ground_truth,workflow ,predicted_label) %>% dplyr::summarise(num_cells = sum(num_cells),percentage =sum(percentage) )  

miscalssifications_plot = ggplot(data = df,aes(y = percentage,fill =  predicted_label , x = workflow)) + 
    theme_bw()+
    geom_bar(stat="identity", position=position_dodge(),color="black", width = 0.7)+
    theme(axis.text.x = element_text(angle=-45, hjust = 0,vjust=1, size = 12),text = element_text(size = 14), 
           plot.title = element_text(size = 14, face = "bold"))+
    ylab("Proportion")+
    xlab("Demultiplexing workflow")+ylim(c(0,max(df$percentage)))+
    scale_fill_discrete(name= "Misclassified as")+ggtitle("Types of miscalssifcations of DU145-tagged cells\n(LMO_custom_cells sample)")

############# per-tag sequencing depth variability
df = data.frame(ground_truth = seurat_object$ground_truth, total_counts = seurat_object$nCount_HTO)
df= df[!df$ground_truth %in% c("Doublet","Negative"),]
df = df %>% dplyr::group_by(ground_truth) %>% dplyr::summarise(seq_depth_range = quantile(total_counts, probs = c(0.99))/
                                                    quantile(total_counts, probs = c(0.01)))

perTag_seqDepthVar = ggplot(data = df,aes(y = seq_depth_range ,x = reorder(ground_truth, -seq_depth_range), fill = ground_truth)) + geom_bar(stat="identity")+ theme_bw()+theme(text = element_text(size=14),legend.position = "None",axis.text.x = element_text(angle=-45, hjust = 0,vjust=1, size = 14))+ xlab("Ground truth label") + ylab("Sequencing depth variability")+
  ggtitle("Per-tag sequencing depth variability \n(LMO_custom_cells sample)")+ theme(plot.margin = margin(0.1,1,2,0.1, "cm"),
                               plot.title = element_text(size = 14, face = "bold")) 

perTag_seqDepthVar
#############
results = seq_depth_effect(seurat_object,tag = tag, predictions = "ground_truth", nbins= 100,norm.method = "CLR_2",ground_truth_column = "ground_truth",pointSize = 0.7)
norm_vs_raw =results$plots[[5]]+ggtitle("Sequencing depth normalized vs raw counts for DU145 tag\n(LMO_custom_cells sample)")+theme(text = element_text(size=14)) +theme(plot.margin = margin(0.1,0.1,4,0.1, "cm"), plot.title = element_text(size = 14, face = "bold")) 

```

## Figure #4

```{r}
ht_fscore = grid.grabExpr(draw(Fscore_heatmap))
ht_dblt= grid.grabExpr(draw(dblt_heatmap))

figure1 <- ggpubr::ggarrange(plotlist = list(ht_fscore,ht_dblt),ncol = 2, nrow = 1,hjust = 0, labels = c("A","B"),label.x = 0)
figure2 <- ggpubr::ggarrange(plotlist = list(perTag_seqDepthVar, miscalssifications_plot, norm_vs_raw),ncol = 3, nrow = 1,hjust = 0, labels = c("C","D","E"),label.x = 0,font.label = list(size = 18))

figure <- ggpubr::ggarrange(plotlist = list(figure1, figure2),ncol = 1, nrow = 2,hjust = 0, heights = c(2,1.5))


# pdf("~/projects/demux-benchmarking/manuscript/Figures/fig_4.pdf", width = 20, height = 15)
# figure
# dev.off()

```

## Comparison of normalization-based workflows with existing tools (simulated datasets)
```{r,message=FALSE}

file = sprintf("%s/simulated.txt",benchmarking_results_dir)
df = read.table(file)
colnames(df) = c( "Dataset","Demultiplexing_method","tag","metric","value")
df = df[ df$metric == "Fscore" & df$tag == "Average" ,]
df$Dataset = stringr::str_remove_all(string = df$Dataset,pattern = ".rep.*")

df = df %>% dplyr::group_by(Dataset,Demultiplexing_method,tag,metric) %>% 
  dplyr::summarise(avg_Fscore = mean(value), std_Fscore = sd(value))

table(df$Dataset)
table(df$Demultiplexing_method)

p1 = ggplot(df, aes(x = reorder(Demultiplexing_method, avg_Fscore), y = avg_Fscore, fill =Demultiplexing_method)) + 
  geom_boxplot(width=0.4,alpha = 1, outlier.shape = NA) + labs(title="",x="Demultiplexing workflows", y = "F score")+ 
  geom_jitter(color="black", size=0.4, width = 0.1) +
  theme_bw()+ 
  theme(text = element_text(size=14),axis.text.x = element_text(angle=0, hjust = 0,vjust=1, size = 14), legend.position = "none", axis.text.y = element_text( size = 14), axis.title.y =  element_text( size = 14))+coord_flip()

#################################### line plot (trend plot)
tmp = stringr::str_split(df$Dataset,pattern = "\\.|_")
df$max.ambient = as.integer(sapply(tmp,FUN = function(x) x[3]))
df$cell_bound = as.integer(sapply(tmp,FUN = function(x) x[5]))
df$cell_bound = paste0( as.character(round(exp(-df$cell_bound) * 100, digits = 0 )), "%")

df$cell_bound = as.factor(df$cell_bound)
df$cell_bound = factor(df$cell_bound, stringr::str_sort(levels(df$cell_bound), numeric = T))

df$max.ambient = as.factor(df$max.ambient)

df2 = df[ grepl(pattern = "Clustering", x = df$Demultiplexing_method),]

p2 = ggplot(df2, aes(x = max.ambient, y = avg_Fscore, color =Demultiplexing_method, group = Demultiplexing_method )) + 
  geom_point(size = 2) +geom_line(size= 1) +facet_wrap(~cell_bound, labeller = label_both, ncol = 2) +
  theme_bw()+
   theme(text = element_text(size=15), axis.text.x = element_text(angle=45, hjust=1, size = 15))+
  ylab("F score")+
  xlab("Maximum ambient contamination")+
   scale_x_discrete(labels= c("100","","500","","900","","1300","","1700",""))+
  ylim(c(0,1))+ scale_color_discrete(name = "Demultiplexing workflows")


figure <- ggpubr::ggarrange(plotlist = list(p1,p2),ncol = 2, nrow = 1,hjust = 0, labels = c("A","B"),label.x = 0.001)
 figure
```

